---
title: "AutoCorrelation Function - ACF"
author: "Dr. Cohen"
format: html
self-contained: true
editor: visual
execute: 
  warning: false
  message: false
---

In practical problems, we do not start with a model, but with observed data
$(x_1, x_2, \ldots , x_n)$. **To assess the degree of dependence** in the data and to select a model for the data, one of the important tools we use is the sample
autocorrelation function (Sample ACF). A few remarks:

-   The sample autocorrelation function (ACF) can be computed for any data
    set and is not restricted to observations from a stationary time series.

-   For data containing a Trend, $|\hat{\rho}(h)|$will display slow decay as h increases.

-   For data containing a substantial deterministic periodic component,
    $|\hat{\rho}(h)|$ will exhibit similar behavior with the same periodicity

```{r}
# Load packages
library(itsmr)
library(tidyverse)
library(tsibble)
library(feasts)
```

## Sample AutoCorrelation Function Examples

### USA population from 1790 - 1990

```{r}

pop = read.table("http://users.stat.umn.edu/~kb/classes/5932/data/uspop.txt", 
                 header = FALSE)
# get data to tsibble format
pop = pop %>% 
  mutate(Year=seq(1790,1990,10), V1=V1/10^6) %>%
  as_tsibble(index = Year) 

# plot the time series
pop %>%
  autoplot(V1) + 
  labs(y="Population in millions")

# plot the ACF
pop %>%
  ACF(V1, lag_max = 20) %>%  
  autoplot()

```

### IID Noise

```{r}

w = rnorm(100,mean=0,sd=1)

w = w %>% 
  as_tibble() %>%
  mutate(t=(1:100)) %>%
  as_tsibble(index = t)

w %>%
  autoplot(value)
  
w %>%
  ACF(value, lag_max = 100) %>% 
  autoplot()

```



### Monthly totals of international airline passengers 1949-1960

```{r}



A = read.table("http://users.stat.umn.edu//~kb/classes/5932/data/airpass.txt", header = FALSE)

A = A %>% 
  as_tibble() %>%
  mutate(mth = rep(yearmonth("1949 Jan")+0:143)) %>%
  as_tsibble()

A %>%
  autoplot(V1) +
  labs(y="Total International of Airline Passengers")
  
A %>%
  ACF(V1, lag_max = 144) %>% 
  autoplot()

```



### Monthly beer production, Australia in Megalitres - Jan. 1956 - Feb 1991

```{r}
B <- read.table("http://users.stat.umn.edu/~kb/classes/5932/data/beer.txt",
                header = FALSE)

B = B %>% 
  select(V3) %>%
  as_tibble() %>%
  mutate(mth = rep(yearmonth("1956 Jan")+0:421)) %>%
  as_tsibble()

B %>%
  autoplot(V3) +
  labs(y="Monthly beer production, Australia in Megalitres")
  
B %>%
  ACF(V3, lag_max = 144) %>% 
  autoplot()


```



## Theoretical ACF for MA(1) Time Series Model

The Moving Average model of order one is given:


$$ 
X_t = \theta Z_{t-1} + Z_t
$$
where $Z_t \sim WN(0, \sigma^2)$

We can show that the ACF $\rho(h)$ of an MA(1) is:

$$
\rho(h) = \theta/(1+\theta^2) \text{ if } \qquad h=+/-1
$$

### Simulate MA(1) data with $\theta=0.8$

```{r}
theta= 0.8
# generate ACF for MA(0.8)
ACF_Theory = ARMAacf(ma =theta, lag.max = 20)
ACF_Theory

# Plot ACF
plot((0:20),ACF_Theory,type="h", xlab = "Lag")

#1. Specify the Model using this function
modelMA1=specify(ma=theta, sigma2 = 1)
#2. generate MA observations with
datasim=sim(modelMA1,n = 100)

# Plot ACF
a=acf(datasim)
a
```


