---
title: "ARIMA Models"
author: "Dr. Cohen"
format: html
self-contained: true
editor: visual
execute: 
  warning: false
  message: false
---

Time Series analysis using AutoRegressive Integrated Moving Average models is appropriate for stationary data. The ARIMA(p,d,q) is given by:

$$
\phi(B) (1-B)^d X_t = \theta(B) Z_t, \qquad Z_t \sim WN(0,\sigma^2)
$$

where $\Phi(B) = 1-\phi_1X_{t-1} - \ldots - \phi_pX_{t-p}$ and $\Theta(B) = 1+\theta_1Z_{t-1} + \ldots + \theta_qZ_{t-q}$

### Quarterly U.S. GNP Gross National Product

```{r}
library(car)
library(ltsa)
library(itsmr)
library(astsa)
library(itsmr)
library(forecast)

# plot
plot(gnp,type="o")
# ACF plot
acf(gnp,lag.max = 100)
# data model
M=c("diff",1)
r = Resid(gnp,M)
test(r)
```

### Test stationary

```{r}
# Augmented Dickey-Fuller Test
# H0: X_t is a non-stationary time series
# H1: X_t is a stationary time series

library(aTSA)
adf.test(r)
# Results in stationary data. No differencing is needed, then d=0.
```

### ARIMA models

```{r}
# p = 1, 2
# q = 1, 2, 3
# ARIMA(2,0,1) == ARMA(2,1)
a21=arima(x = r,order = c(2,0,1))
test(a21$residuals)

# ARIMA(1,0,1) == ARMA(1,1)
a11=arima(x = r,order = c(1,0,1))
test(a11$residuals)

# ARIMA(2,0,0) == ARMA(2,0)
a2=arima(x = r,order = c(2,0,0))
test(a2$residuals)

#abest=autofit(r,p = 0:2,q=0:2)
#test(abest)
```

We will try a transformation to reduce the effect of variance change.

```{r}
l=powerTransform(gnp)
l$roundlam
## we can use log() transformation

plot(log(gnp),type="o")
acf(log(gnp),lag.max = 200)
# data model
M=c("log","diff",1)
r = Resid(gnp,M)
test(r)
# Stationary
adf.test(r) # pvalue <0.05 => data is stationary


# ARIMA(1,0,1) == ARMA(1,1)
arma11=arima(x = r,order = c(1,0,1))
test(arma11$residuals)

# ARIMA(2,0,1) == ARMA(2,1)
arma21=arima(x = r,order = c(2,0,1))
test(arma21$residuals)

# ARIMA(2,0,0) == ARMA(2,0) == AR(2)
ar2=arima(x = r,order = c(2,0,0))
test(ar2$residuals)

# ARIMA(1,0,2) == ARMA(1,2)
arma12=arima(x = r,order = c(1,0,2))
test(arma12$residuals)


# ARIMA(0,0,2) == MA(2)
ma2=arima(x = r,order = c(0,0,2))
test(ma2$residuals)


## Comparing AICs
c(MA2=ma2$aic, ARMA12=arma12$aic, ARMA21=arma21$aic)

```

The equation of MA(2):

$$
(1-B)\log(X_t) = 0.3Z_{t-1} + 0.2Z_{t-2} + Z_t, \qquad Z_t \sim WN(0,8.919e-05)
$$

### Example Quarterly Time Series of the Number of Australian Residents

```{r}

plot.ts(austres, type="o")
plota(austres)
     

M=c("diff",1)
e = Resid(austres,M)
test(e)

aTSA::adf.test(e)
# since the non-stationary was not rejected. We need to do another differencing operator.

aTSA::adf.test(diff(e))

test(diff(e))

# Fit ARIMA(p,d=1,q)
a1=arima(x = austres,order = c(2,1,2))
test(a1$residuals)

# Fit ARIMA(p,d=2,q)
a2=arima(x = austres,order = c(2,2,2))
test(a2$residuals)


# Fit ARIMA(p,d=1,q) - not iid
a3=arima(x = austres,order = c(1,1,0))
test(a3$residuals)


# Fit ARIMA(p,d=1,q) - not iid
a4=arima(x = austres,order = c(1,1,1))
test(a4$residuals)


# Fit ARIMA(p,d=2,q) - not iid
a5=arima(x = austres,order = c(1,2,0))
test(a5$residuals)


# Fit ARIMA(p,d=2,q) 
a6=arima(x = austres,order = c(1,2,1))
test(a6$residuals)


# Fit ARIMA(p,d=2,q) = IMA(2,1)
a7=arima(x = austres,order = c(0,2,1))
test(a7$residuals)

c(a1$aic,a2$aic,a6$aic,a7$aic)

```

The equation of ARIMA(2,1,2):

$$
(1-0.6843B-0.3121B^2)(1-B)(X_t) = (1 - 0.1961B  -0.2975B^2)Z_t, \qquad Z_t \sim WN(0,99.47)
$$

The equation of IMA(2,1):

$$
(1-B)^2 X_t = -0.59Z_{t-1} + Z_t; Z_t \sim WN(0,\sigma^2 = 101.17)
$$
