---
title: "ARMA Modeling"
author: "Dr. Cohen"
format: html
self-contained: true
editor: visual
execute: 
  warning: false
  message: false
---

Time Series analysis using ARMA models can be done as follows:

1.  Plot the data / ACF and examine the characteristics

2.  Data model (variance change, trend, seasonality)

3.  Transform and Eliminate / Estimate the components

4.  Test the residuals if IID

5.  Select p and q for the ARMA(p,q) using PCAF and ACF

6.  Select the best

## Estimation of the parameters

ARMA(p,q) are stationary time series models given by:

$$
\Phi(B) X_t = \Theta(B) Z_t, \qquad Z_t \sim WN(0,\sigma^2)
$$

where $\Phi(B) = 1-\phi_1X_{t-1} - \ldots - \phi_pX_{t-p}$ and $\Theta(B) = 1+\theta_1Z_{t-1} + \ldots + \theta_qZ_{t-q}$

### Simulate ARMA models

#### MA(1) example

$$
X_t = 0.8Z_{t-1} + Z_t;  Z_t \sim WN(0,1)
$$

```{r}
library(itsmr)
ma1 = arima.sim(model = list(ar=c(),ma=c(0.8)),n=200)
plot.ts(ma1)
plota(ma1)

# Innovations estimator - MA(q)
ia(ma1,q=1)
```

#### AR(1) example

$$
X_t = 0.4 X_{t-1} + Z_t; Z_t \sim WN(0,1)
$$

```{r}
library(itsmr)
ar1 = arima.sim(model = list(ar=c(0.4),ma=c()),n=200)
plot.ts(ar1)
plota(ar1)

# yule-walker estimator - AR(p)
yw(ar1,p=1)
```

#### ARMA(1,1) example

$$
X_t = 0.4X_{t-1} + 0.5 Z_{t-1} + Z_t; Z_t \sim WN(0,1)
$$

```{r}
library(itsmr)
arma11 = arima.sim(model = list(ar=c(0.4),ma=c(0.5)),n=200)
plot.ts(arma11)
plota(arma11)

# HR estimator - ARMA(p,q)
hannan(arma11,p = 1,q = 1)
```

### MLE estimator

```{r}
library(itsmr)
arma11 = specify(ar=c(0.4),ma=c(0.5),sigma2=1)
check(arma11)
data.sim = sim(arma11,n=200)
plot.ts(data.sim)
acf(data.sim,lag.max = 40)

# MLE estimator - ARMA(p,q)
arma(data.sim,p=1,q=1)
```

## Example Dow Jones Utilities Index 1972

```{r}
plot.ts(dowj)
acf(dowj,lag.max = 50)
```

Our data model:

$X_t = m_t + Y_t$

```{r}
# differencing operator
r = diff(dowj)
mean(r)
R = r - mean(r)
test(R)
```

Some possible models are: - AR(1) or MA(1) or MA(2) - ARMA(1,1) or ARMA(1,2)

```{r}
# AR(1)
ar1 = arma(x = R,p = 1,q=0)
ma1 = arma(x = R,p = 0,q=1)
ma2 = arma(x = R,p = 0,q=2)
arma11 = arma(x = R,p = 1,q=1)
arma12 = arma(x = R,p = 1,q=2)
# autofit
automodel = autofit(R)

cbind(ar1=ar1$aicc,
      ma1=ma1$aicc,
      ma2=ma2$aicc,
      arma11=arma11$aicc,
      arma12=arma12$aicc,
      automodel=automodel$aicc)
```

The best model in terms of AICC is AR(1) given by:

$$
 (1-0.45B)(\nabla X_t - 0.134) = Z_t, \qquad Z_t \sim WN(0,0.145) 
$$

Or we can rewrite it: $X_t - X_{t-1} - 0.134= Y_t$

$$
Y_t = 0.45Y_{t-1} + Z_t
$$

We need to see if the the residuals of the model are iid?

```{r}
model = specify(ar=arma11$phi , ma=arma11$theta , sigma2 = arma11$sigma2)
M= c("diff",1)
rfit = Resid(x = dowj,M = M,a = model) # residual from the model Z_t
test(rfit)
```

The AR(1) model is not adequate since the iid hypothesis was rejected.

We fit ARMA(1,1) with AICC=75.13 and the model is adequate:

The equation is:

$$
(1-0.77B)(\nabla X_t - 0.134) = Z_t -0.42Z_{t-1}, \quad Z_t \sim WN(0,0.142)
$$ The 95% CI for the parameters are:

```{r}
c(arma11$phi + 1.95 *arma11$se.phi,arma11$phi - 1.95* arma11$se.phi)
c(arma11$theta + 1.95 *arma11$se.theta,arma11$theta - 1.95 *arma11$se.theta)
```

## Wine data

```{r}
plot.ts(wine)
acf(wine)
```

We do have: trend (line) and seasonality (d=12)

```{r}
data.model = c("season",12,"trend",1)
r = Resid(wine,M=data.model) # yt = xt - mt - st - 1.347
test(r)
# autofit
bestmodel = autofit(r,p=0:2,q=0:2)
# test iid for the fit
model = specify(ar=bestmodel$phi , ma=bestmodel$theta , sigma2 = bestmodel$sigma2)
rfit = Resid(x = wine,M = data.model,a = model)
test(rfit)
```

We write the model:

$$
 X_t = 1.347 + m_t + s_t + 0.18X_{t-1} + 0.19X_{t-2} + Z_t, \quad Z_t \sim WN(0,30746.27)
$$
