---
title: "GARCH Models"
author: "Dr. Cohen"
format: html
self-contained: true
editor: visual
execute: 
  warning: false
  message: false
---

This lecture will cover:

-   The ARCH (autoregressive conditional heteroscedasticity) model
-   The GARCH model (Generalized ARCH)

Financial time series data are special because of their features, which include tail heaviness, asymmetry, volatility, and serial dependence without correlation.

Let consider $P_t$ the price of a stock of other financial asset at time $t$, then we can define the log return by $Z_t = log(P_t) − log(P_{t−1})$ A model of the return series should support the fact that this data has a conditional variance $h_t$ of $Z_t$ is not independent of past value of $Z_t$.

The idea of the ARCH model is to incorporate the sequence $h_t$ in the model by:

$$
Z_t =\sqrt{h_t e_t}, \qquad e_t \sim IIDN(0, 1)
$$

where $h_t$ is known as the **volatility** and related to the past values of \$Z_t\$ via the **ARCH(p)** model:

$$
h_t = \alpha_0 + \sum_{i=1}^{p} \alpha_i Z^2_{t-i} 
$$

The **GARCH(p,q)** model is given by:

$$
h_t = \alpha_0 + \sum_{i=1}^{p} \alpha_i Z^2_{t-i} + \sum_{j=1}^{q} \beta_j h_{t-j}
$$

$\alpha_0 > 0 \text{ and } \alpha_i ≥ 0, \beta_j ≥ 0$

### Google Share Price

```{r}
library(rugarch) # run GARCH models
library(quantmod) # upload financial data from Yahoo Finance
library(itsmr)
library(aTSA)

getSymbols("GOOG", from = as.Date("2015-01-01"), to = as.Date("2023-3-30"))

# Select Open Price
op=GOOG$GOOG.Open
# plot Time Series
plot(op)

# Return data 
R= diff(log(as.numeric(op)))

# Plot Return
plot(R)

# Plot ACF 
acf(R,lag.max = 50)

# Plot ACF of R^2
acf(R^2, lag.max = 50)
```

## GARCH model

```{r}

# Specify the model
spec1 = ugarchspec()
spec1

# Mean Model  AR(1)
spec1 = ugarchspec(mean.model = list(armaOrder=c(1,0)))
spec1

# Mean Model- EWMA
spec2 = ugarchspec(mean.model = list(armaOrder=c(0,0)),variance.model = list(garchOrder =c(2,0)))
spec2

# Fit GARCH
ugfit = ugarchfit(spec = spec2, data = R)
ugfit

# Forecast with GARCH
ugf = ugarchforecast(ugfit, n.ahead = 10)
ugf

plot(ugf@forecast$sigmaFor,type="l")
plot(ugf@forecast$seriesFor,type="l")


```

###  NYSE data

```{r}
library(astsa)

plot(nyse,type="o")

plota(nyse)

plota(nyse^2)

spec2 = ugarchspec(mean.model = list(armaOrder=c(1,0)),

                   variance.model = list(garchOrder =c(1,1)))

spec2

# Fit GARCH

ugfit = ugarchfit(spec = spec2, data = nyse)

ugfit

# Forecast with GARCH

ugf = ugarchforecast(ugfit, n.ahead = 10)

ugf

plot(ugf@forecast$sigmaFor,type="l")

plot(ugf@forecast$seriesFor,type="l")
```
