---
title: "HW2 Keys Time Series"
author: "Dr. Cohen"
format: html
self-contained: true
editor: visual
execute: 
  warning: false
  message: false
---

```{r}
# Load packages
library(itsmr)
library(tidyverse)
library(tsibble)
library(feasts)
```


# Number of international airline passengers, 1949 to 1960. 

Plot the data: `airpass{itsmr}`

```{r}
# get data to tsibble format
airpassts = airpass %>% 
  as_tibble() %>%
  mutate(mth=seq(yearmonth("1949 Jan"),yearmonth("1960 Dec"),1)) %>%
  as_tsibble() 

# plot the time series
airpassts %>%
  autoplot(value) + 
  labs(y="Number of airline passengers",
       title="Number of international airline passengers 1949 to 1960.")

# plot the ACF
airpassts %>%
  ACF(value, lag_max = 140) %>%  
  autoplot()
```
The series shows a upward trend and a season of 12 with variance change. I shall consider the following model:

$$ X_t = m_t + s_t + Y_t $$ where

-   $m_t$ is the trend component

-   $s_t$ is the seasonal component

-   $Y_t$ random noise component / Residuals

## Transformation
```{r}
library(car)
L = powerTransform(airpassts$value)
L$roundlam

airpassts %>%
  mutate(valueT = log(value)) ->airpassts
```
the $\log$ transformation is appropriate to use. 

## Trend Estimation with MA smoothing

```{r}


airpassts %>%
  mutate(ma.trend3 = smooth.ma(valueT, 3),
         ma.trend5 = smooth.ma(valueT, 5),
         ma.trend7 = smooth.ma(valueT, 7),
         ma.trend9 = smooth.ma(valueT, 9)) %>% # q=3 means averaging 2q+1 values
  pivot_longer(cols =-c(mth,value),
               names_to = "series",
               values_to = "values") %>%
    group_by(series) %>%
    autoplot(values)


airpassts = airpassts %>%
  mutate(ma.trend7 = smooth.ma(valueT, 7))
```

The trend with $q=7$ or $q=9$ seems to be reasonable.


#### Trend Estimation: Regression

```{r}

airpassts %>%
  mutate(reg.trend1 = trend(valueT, 1),# 1 means polynmial of degree 1 
         reg.trend2 = trend(valueT, 2)) %>% 
  pivot_longer(cols =-c(mth,value),
               names_to = "series",
               values_to = "values") %>%
    group_by(series) %>%
    autoplot(values)

airpassts = airpassts %>%
  mutate(reg.trend1 = trend(valueT, 1))
```

I will consider a straight line trend.

## Seasonality Estimation

```{r}
airpassts = airpassts %>%
  mutate(season.value=season(valueT, d=12))
```

#### Plots the series and trend estimates
```{r}
airpassts %>%
  pivot_longer(cols =-c(mth,value,season.value),
               names_to = "series",
               values_to = "values") %>%
    group_by(series) %>%
    autoplot(values)

airpassts = airpassts %>%
  mutate(reg.trend1 = trend(valueT, 1))
```


#### Test IID hypothesis for the residuals

Use the`Resid()` function to find the residuals.

```{r}
# deaths
data.model = c("log","season",12,"trend",1) # from examining the series
R= Resid(airpass,data.model)
# Test IDD
test(R) # need to tests to fail to reject the null hypothesis

```

Further analysis is needed because the iid hypothesis was not rejected.

### Trend and Seasonality Elimination using the Differencing operator 

$$
Y_t = \nabla \nabla_{12} \log(X_t)
$$

```{r}
R = diff(diff(log(airpass),lag = 12,differences = 1))
test(R)
```

Further analysis is needed because the IID hypothesis was not rejected.


## Level of Lake Huron, 1875 to 1972.

lake{itsmr}: Level


```{r}
# get data to tsibble format
lakets = lake %>% 
  as_tibble() %>%
  mutate(year=1875:1972) %>%
  as_tsibble(index=year) 

# plot the time series
lakets %>%
  autoplot(value) + 
  labs(y="Number of airline passengers",
       title="Number of international airline passengers 1949 to 1960.")

# plot the ACF
lakets %>%
  ACF(value, lag_max = 140) %>%  
  autoplot()
```
